% =====================================================================
%  Chapter 1 --- General Context of the Project
% =====================================================================
\chapter{General Context of the Project}
\label{chap:context}

\section{Introduction}
The objective of this chapter is to situate the present graduation project in its industrial and academic context. The chapter introduces the host organization and its existing operational ecosystem, describes in detail the three databases on which the company currently relies, identifies the structural limitations that prevent the company from extracting analytical value from its historical telemetry, formalizes the problem statement and the project objectives, sketches the proposed solution and its data warehousing strategy, and concludes by comparing three candidate project-management methodologies before justifying the adoption of CRISP-DM. The chapter therefore establishes the foundations on which all subsequent technical work --- data understanding, preparation, modeling and deployment --- is grounded.

\section{Industrial and academic context}
The economic relevance of fleet analytics has grown steadily over the past decade. The combined drop in the price of GPS trackers, the densification of mobile coverage and the emergence of cloud-native data platforms have made it economically viable for medium-sized fleet operators to instrument every vehicle and to retain the resulting telemetry. The competitive advantage no longer lies in collecting data --- which has effectively become a commodity --- but in \emph{transforming} it into reliable, low-latency analytical signals that support faster, evidence-driven decisions.

In parallel, academic research on industrial data science has converged on a small set of structured methodologies that align project-management practice with the iterative nature of analytics. Among them, CRISP-DM has remained the most adopted framework for two complementary reasons: it is vendor-neutral, and its six phases (business understanding, data understanding, data preparation, modeling, evaluation, deployment) match the natural decomposition of a graduation project of this kind. The present work explicitly adopts this methodology and uses it as the structuring principle of the report.

\section{Host organization}

\subsection{Presentation of \hostorg}
\hostorg{} is a Tunisian operator of a multi-tenant telematics platform. The company designs, integrates and operates GPS tracking solutions for logistics fleets, distributing physical devices, ingesting their telemetry through SIM-card connectivity, and offering a software portal through which fleet managers monitor their vehicles in real time. The platform supports several customer fleets, denoted \emph{tenants}, each one comprising tens to hundreds of vehicles. Each tenant has its own subgroup of tables and identifiers in the company's databases, reflecting the contractual and operational separation between customers.

\subsection{Activities and business areas}
The activities of \hostorg{} cover four complementary segments:

\begin{itemize}
    \item the integration of GPS hardware on customer vehicles and the management of the SIM-card connectivity required for telemetry uplink;
    \item the operation of a multi-tenant SaaS platform exposing live tracking, geofencing, alerts and standard reports;
    \item the production of operational and managerial reports for fleet managers (mileage, fuel consumption, alerts, maintenance);
    \item the exploitation of the historical archive for value-added analytical use cases, of which the present project is a representative example.
\end{itemize}

\subsection{Organizational structure of the internship}
The internship has been hosted in the technical division of \hostorg, working closely with the data engineering and software engineering teams. The project has been conducted by a binomial of two student engineers who collaborated remotely from separate workstations. To make this collaboration possible while preserving the confidentiality of the company's data, an analytical infrastructure dedicated to the project has been provisioned on a Microsoft Azure virtual machine; access to that infrastructure has been mediated by SSH tunnelling, with a strict IP allowlist restricted to the two students' workstations and to the office network of \hostorg.

\section{Analysis of the existing system}

\subsection{Operational landscape of \hostorg}
The operational platform of \hostorg{} relies on three databases, each fulfilling a distinct role in the lifecycle of a telematics signal --- from its arrival on the wire to its consumption by the customer-facing portal. The database management system (DBMS) historically retained by the company is \emph{MySQL}; it carries the entire production workload and constitutes the single source of truth from which the present project derives its data. The three databases are presented below in the order in which a telemetry frame traverses them.

\subsubsection{Database 1 --- raw telemetry stream}
The first database is the entry point of the telematics signal. It receives the frames emitted by the GPS and IoT devices installed on the vehicles, transmitted over the cellular network through SIM cards. The frames are persisted as they arrive, in their native binary representation; in the database, they appear as \emph{hexadecimal} payloads stored in dedicated columns of the ingestion tables. The ingestion is performed by internal collectors that play a role functionally analogous to streaming consumers (in the spirit of a Kafka consumer), translating the device protocol into a row-oriented persistence layer. Because the payloads are not yet decoded at this stage, they are not directly queryable through standard SQL idioms; their analytical value can only be unlocked once they have been parsed.

\subsubsection{Database 2 --- decoded archive}
The second database hosts the \emph{decoded} archive of the telemetry. It is populated by an internal toolchain that reads the hexadecimal payloads of the first database and transforms them into a readable, exploitable representation: timestamps, latitude--longitude pairs, instantaneous speed, ignition state, accelerometer values, RPM, fuel level, engine status and the other physical channels emitted by the device. The decoded rows are stored in tables whose names follow a recognizable convention of the form \texttt{arch\_xxxxx}, the suffix \texttt{xxxxx} encoding either the tenant or a logical partition of the archive. The archive retains approximately the most recent six months of data; older slices are evicted by a retention process. This database is the principal source of historical telemetry exploited by the present project.

\subsubsection{Database 3 --- users, clients and master data}
The third database carries the master data of the platform: customers, users, drivers, vehicles, contracts, and the cross-references that bind a fleet to its tenant. Its tables follow the convention \texttt{user\_xxxxx} and are organized by tenant subgroup, reflecting the multi-tenant nature of the SaaS offering. The tenant identifier, propagated throughout the schema, is the cornerstone of the partitioning logic: every operational query, every report and every analytical workload restricts its scope to the tenant of interest, both for confidentiality reasons and for analytical correctness.

\subsection{Decoding pipeline and data flow}
The three databases are chained by a decoding pipeline that transforms the hexadecimal payloads of the first database into the readable rows of the second. The decoding scripts implement the device protocol, validate the structural integrity of each frame, derive higher-level signals (for example, the per-trip aggregates of distance, duration and maximum speed) and write the resulting rows to the \texttt{arch\_xxxxx} tables. Master data from the third database is then joined at query time to attach a vehicle, a driver, a tenant and the relevant contractual context to each row. This three-stage flow --- raw hex, decoded archive, master data --- is the existing data architecture of \hostorg{} on which the present project is grafted.

\subsection{Analytical surface of the existing system}
On top of the three databases, the company exposes a set of dashboards and standard reports. These deliverables address well-established descriptive needs: mileage per vehicle, alert counts, daily activity, fuel consumption, simple comparisons between months. They serve as the day-to-day instrumentation of fleet managers and their value is not in question. They do, however, share a common limitation: their semantics are descriptive, not predictive; they answer \emph{what happened} but not \emph{which devices are likely to behave abnormally}, \emph{how can the fleet be segmented into operational profiles}, or \emph{which vehicles should be prioritized for inspection}.

\subsection{Identified limitations}
A diagnostic of the existing analytical surface, conducted at the start of the internship together with the engineering team of \hostorg, surfaced six structural limitations that motivate the present project:

\begin{enumerate}
    \item \emph{Descriptive scope}: the dashboards summarize past activity but expose no predictive or prescriptive layer.
    \item \emph{Under-exploitation of the archive}: the six months of decoded telemetry stored in the \texttt{arch\_xxxxx} tables represent a substantial historical asset that is queried only marginally beyond simple aggregations.
    \item \emph{Absence of a behavioural model}: there is no segmentation of the devices into operational profiles, no anomaly detection layer, and no notion of a continuous risk score that would prioritize the operations team's attention.
    \item \emph{Tight coupling between operational storage and analytics}: the operational MySQL databases are designed for transactional access, not for analytical workloads; running heavy aggregations on them risks competing with the live customer-facing traffic.
    \item \emph{Limited collaboration surface}: the internal databases are reachable only from inside the company network, which is a sensible operational decision but complicates remote development by a binomial of student engineers working on personal workstations.
    \item \emph{Limited observability}: there is no run log, no quality check and no drift indicator that would let an operator certify the daily health of any analytical layer built on top of the data.
\end{enumerate}

\subsection{Synthesis of the diagnostic}
The existing system therefore fulfils its descriptive mission but constitutes a structural obstacle to the deployment of any predictive analytical use case --- and especially to the \emph{Device Behavior Scoring \& Risk Classification} workstream that is the explicit objective of the present internship. The challenge is twofold: extract analytical value from the archive without disturbing the operational MySQL databases, and host the analytical work in an environment that the binomial can operate jointly under proper security guarantees.

\section{Problem statement}
The starting observation of this thesis is that \hostorg{} sits on a high-volume telemetry archive whose analytical potential remains largely untapped. The company can already \emph{observe} its fleet through dashboards and standard reports, but it cannot yet \emph{anticipate} the operational risks carried by individual devices, segment the fleet into actionable behavioural profiles or prioritize the interventions of its operations team on the basis of an objective, data-driven score. The problem is not visualization in itself --- the existing dashboards are functional --- but the absence of an advanced analytical layer that would transform the raw and archived telemetry into forward-looking, decision-grade signals.

The problem addressed in this thesis can therefore be stated as follows. \emph{Given the three operational MySQL databases of \hostorg, namely the raw hexadecimal telemetry database, the decoded \texttt{arch\_xxxxx} archive and the \texttt{user\_xxxxx} master data, and given the practical constraint that a binomial of student engineers must collaborate remotely on the project, how can the historical telemetry be repatriated, cleaned and modelled into a transparent, incremental analytical platform that produces an interpretable behavioural segmentation and an explicit risk score per device-month, while preserving the integrity of the operational sources and the confidentiality of the data?}

The reformulation emphasizes four entangled objectives: (i) to migrate the relevant slices of the operational MySQL data into an analytical environment without polluting the production workload; (ii) to formalize a transparent and auditable cleaning layer that reflects the physical reality of telemetry; (iii) to produce a per-device, per-month behavioural representation amenable to unsupervised modelling; and (iv) to expose the analytical outputs to the consumers of the host organization through an API and a dashboard.

\section{Proposed solution}

\subsection{Functional outline}
The proposed solution consists of an end-to-end CRISP-DM-aligned analytical platform deployed on Microsoft Azure. Slices of the three operational databases of \hostorg{} are exported as MySQL dump files, converted to a PostgreSQL-compatible representation by dedicated scripts written for the project, and loaded into a bronze \emph{staging} schema of an analytical PostgreSQL instance. A transparent eleven-rule cleaning engine then materializes a silver \emph{warehouse} schema composed of five dimensions and eleven facts, on top of which a gold \emph{marts} schema exposes the device-month behavioural mart and a family of business-intelligence views. A per-tenant K-Means segmentation and an Isolation Forest risk score are trained on the marts; their outputs are persisted, served through a Python API and published to an interactive dashboard.

\subsection{Why an analytical PostgreSQL warehouse?}
The strategic decision of the project is to keep the operational MySQL databases of \hostorg{} untouched and to constitute, alongside them, a dedicated analytical warehouse on PostgreSQL~16. PostgreSQL has been selected for six independent reasons. \emph{First}, it offers strong support for analytical workloads, with mature window functions, common table expressions, lateral joins and partitioning capabilities that significantly simplify the encoding of behavioural indicators. \emph{Second}, it is fully ACID-compliant, which is a sine qua non for an industrial pipeline that runs unattended every five minutes. \emph{Third}, its indexing layer (B-tree, BRIN, GIN, partial and expression indexes) covers a broader range of access patterns than the operational MySQL setup of the company. \emph{Fourth}, it scales gracefully to the volumes considered by the project, namely several tens of millions of telemetry rows. \emph{Fifth}, it is highly extensible, with a rich ecosystem of extensions (\texttt{pg\_stat\_statements}, \texttt{pgcrypto}, \texttt{postgis}, \texttt{pg\_partman}, etc.) that opens future trajectories such as geospatial analytics or fine-grained workload telemetry. \emph{Sixth}, it is a first-class citizen of the modern data and BI ecosystem, with native or near-native compatibility with the Python data stack (\texttt{psycopg2}, SQLAlchemy, pandas), with orchestration platforms and with standard BI tools, which will matter when the analytical results are published to fleet managers.

\subsection{Why a Microsoft Azure virtual machine?}
The analytical PostgreSQL instance is hosted on a Microsoft Azure virtual machine for two complementary reasons. The first one is technical: Azure offers a controlled and reproducible environment in which the storage, the compute and the networking can be provisioned through code and recovered after an incident. The second one is collaborative: because the project is conducted by a binomial of student engineers working on separate personal machines, the database cannot reasonably be hosted on a single laptop; it must live in a shared, always-available environment. The Azure VM serves precisely this purpose. Access to the VM is mediated by SSH tunnelling, with a strict IP allowlist enforced at the Azure Network Security Group level: only the two students' workstations and the office network of \hostorg{} can reach the SSH port, and the PostgreSQL port itself is never exposed to the public internet.

\subsection{Architectural outline}
At the architectural level, the solution adopts a three-layer warehouse pattern (staging, warehouse, marts), a watermark-driven incremental processing model, a Prefect orchestration of the batch flow, a five-minute cron schedule on the Azure VM and an explicit monitoring layer. The MySQL-to-PostgreSQL conversion bridge, the SSH tunnel for shared access and the IP-restricted Network Security Group complete the architectural blueprint that is detailed in Chapter~\ref{chap:business}.

\subsection{Expected benefits}
The expected benefits, measured against the limitations of the existing system, are: (i) the unlocking of the analytical value carried by the decoded archive of \hostorg, without disturbing the operational MySQL workload; (ii) the introduction of a transparent and auditable cleaning rule engine that survives schema evolution and operator turnover; (iii) the availability of a versioned, declarative feature contract suitable for behavioural segmentation and for risk scoring; (iv) the coverage of the pipeline by an explicit run log, a layered data quality monitoring and a model drift indicator; and (v) a stable consumption surface (marts and views) for the BI and the API consumers of the platform.

\section{Project management methodology}

\subsection{Comparison of candidate methodologies}
Three candidate methodologies have been compared at the start of the project: CRISP-DM, GIMSI and Scrum. The first one is a domain-specific data-mining methodology, the second is a methodology for the design of management dashboards, and the third is a generic agile software-development framework. Table~\ref{tab:method-compare} summarizes the comparison along five criteria: scope, granularity of the phases, alignment with data science deliverables, support for iterative refinement, and operational maturity in the industrial setting of \hostorg.

\begin{table}[h]
\centering
\caption{Comparison of the candidate project-management methodologies.}
\label{tab:method-compare}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Criterion} & \textbf{CRISP-DM} & \textbf{GIMSI} & \textbf{Scrum} \\
\hline
Domain orientation & Data mining & Management dashboards & Generic software \\
Phase decomposition & 6 phases & 10 steps & Sprints, no phases \\
Data science alignment & Strong & Partial & Weak \\
Iterative refinement & Native & Limited & Native \\
Industrial maturity & High & Medium & High \\
\hline
\end{tabular}
\end{table}

\subsection{Justification of the choice of CRISP-DM}
CRISP-DM has been retained for three reasons. First, its six phases match the natural decomposition of the project deliverables (business understanding, data, preparation, modeling, evaluation, deployment), avoiding any artificial mapping between the methodology and the work. Second, CRISP-DM places explicit weight on the \emph{business understanding} and the \emph{evaluation} phases, both of which are critical in an industrial context where misalignment between the model and the operational reality is the principal cause of failure. Third, CRISP-DM coexists peacefully with agile software practices for the engineering bricks (the orchestration, the API, the dashboard), in the sense that each CRISP-DM phase can be delivered through one or several Scrum-flavoured iterations without conflict.

\subsection{Project plan along the six CRISP-DM phases}
The project plan has been organized as a sequence of six CRISP-DM phases, with explicit deliverables at the end of each one: a business understanding document and a target architecture (Phase~1), a data understanding report on the three operational databases of \hostorg{} (Phase~2), a curated PostgreSQL warehouse and a feature mart populated from the converted MySQL dumps (Phase~3), a trained per-tenant K-Means and Isolation Forest model (Phase~4), an evaluation report with internal metrics and stability proxies (Phase~5) and an industrialized pipeline with API, dashboard and monitoring on Azure (Phase~6).

\section{Conclusion}
This first chapter has positioned the project in its industrial and academic context, presented \hostorg{} and its three operational MySQL databases, characterized the existing analytical surface and its structural limitations, formulated a problem statement centred on the under-exploitation of the historical telemetry archive, motivated the strategic choice of a dedicated PostgreSQL warehouse hosted on a Microsoft Azure virtual machine for the binomial of student engineers, and justified the adoption of CRISP-DM as the project-management methodology. The next chapter operationalizes this strategic outline by formalizing the business understanding, the research questions, the requirements, the target architecture and the technology stack of the solution.
