% =====================================================================
%  Chapter 4 --- Data Preparation
% =====================================================================
\chapter{Data Preparation: Conversion, ETL and Feature Engineering}
\label{chap:prep}

\section{Introduction}
Building on the data understanding of Chapter~\ref{chap:data}, the present chapter formalizes the data preparation phase of CRISP-DM. The chapter organizes the work into seven sections: (i) the conversion bridge that transforms the MySQL dump files of \hostorg{} into a PostgreSQL-importable representation; (ii) the construction of the ETL pipeline; (iii) the eleven-rule cleaning engine; (iv) the dimensional modeling of the warehouse; (v) the materialization of the analytical marts; (vi) the engineering of the feature registry; and (vii) the validation of the prepared dataset. Each section is illustrated by SQL or Python excerpts, the full versions of which are listed in the appendices.

\section{From MySQL dumps to a PostgreSQL staging layer}

\subsection{Rationale and scope}
The first preparation step is the conversion of the MySQL dump files received from \hostorg{} into a representation that PostgreSQL can ingest reliably. This step is necessary because the operational databases of \hostorg{} run on MySQL while the analytical warehouse of the project runs on PostgreSQL~16. The conversion bridge is therefore not an accessory utility: it is the technical interface between the operational ecosystem of the company and the analytical platform built for the project.

\subsection{Dialect-level translation}
A native MySQL dump cannot be replayed verbatim into PostgreSQL because the two dialects diverge on several points. The bridge addresses those divergences explicitly:

\begin{itemize}
    \item \emph{Type translation}: MySQL types such as \texttt{TINYINT(1)}, \texttt{DATETIME}, \texttt{TEXT}, \texttt{BLOB} and \texttt{ENUM} are mapped to their PostgreSQL counterparts (\texttt{boolean}, \texttt{timestamp}, \texttt{text}, \texttt{bytea}, \texttt{text} with a \texttt{CHECK} constraint).
    \item \emph{Identifier and string quoting}: MySQL backticks (\texttt{`column`}) are replaced by PostgreSQL double quotes; embedded single quotes inside string literals are normalized to the PostgreSQL escape convention.
    \item \emph{Auto-increment columns}: MySQL \texttt{AUTO\_INCREMENT} columns are converted to PostgreSQL \texttt{GENERATED BY DEFAULT AS IDENTITY} columns, preserving the original sequence values when present.
    \item \emph{Default expressions}: MySQL-specific defaults (\texttt{ON UPDATE CURRENT\_TIMESTAMP}, \texttt{0000-00-00} sentinels) are rewritten or nullified to match PostgreSQL semantics.
    \item \emph{Character-set normalization}: the dumps are normalized to UTF-8 (NFC), which is the default encoding of the analytical PostgreSQL instance.
    \item \emph{Index and constraint translation}: MySQL-specific clauses (\texttt{ENGINE=InnoDB}, \texttt{COLLATE}, \texttt{KEY ... USING BTREE}) are stripped or rewritten so that the resulting DDL is accepted by PostgreSQL.
\end{itemize}

\subsection{Conversion pipeline}
The conversion bridge has been implemented as a small Python toolchain documented in Appendix~\ref{app:python}. For each dump file, the toolchain (i) parses the MySQL DDL and re-emits an equivalent PostgreSQL DDL; (ii) extracts the \texttt{INSERT} statements and rewrites them as PostgreSQL \texttt{COPY} payloads, which are an order of magnitude faster to load on large tables; (iii) writes the converted artefact under \texttt{exports/}; and (iv) loads it into the staging schema of the analytical PostgreSQL through a single transaction. The bridge is idempotent: re-running it on the same dump produces the same staging state.

\subsection{Hexadecimal payloads of the raw telemetry database}
As described in Chapter~\ref{chap:context}, the first operational database of \hostorg{} stores the raw telemetry as hexadecimal payloads. The internal decoding scripts of the company already transform those payloads into the readable rows of the \texttt{arch\_xxxxx} archive, and the present project consumes the result of this decoding rather than re-implementing it. The conversion bridge therefore does not need to handle hexadecimal payloads directly, except as opaque \texttt{bytea} columns when they incidentally appear in the dumps; the analytical-grade telemetry signals enter the staging layer in their decoded form.

\section{Pipeline architecture}

\subsection{Pipeline stages}
Once the staging layer is populated by the conversion bridge, the analytical pipeline is decomposed into seven sequential stages, each idempotent: \emph{extraction} from the source MySQL dumps to the converted artefacts; \emph{loading} into staging; \emph{cleaning} of the staging slice; \emph{loading} of the dimensions; \emph{loading} of the facts; \emph{materialization} of the marts and views; and \emph{validation} of the prepared dataset.

\subsection{Watermark-driven incremental processing}
The downstream stages of the pipeline (after the staging load) are driven by a watermark stored in \texttt{warehouse.etl\_watermark}. Each cycle reads the latest watermark per source--target pair, processes only the slice strictly after that watermark (with a configurable overlap), and updates the watermark at the end of the successful run. Listing~\ref{lst:watermark} shows a representative implementation.

\begin{lstlisting}[style=sql,caption={Reading and updating a watermark.},label={lst:watermark}]
-- Read the watermark
SELECT last_event_ts
FROM warehouse.etl_watermark
WHERE source_table = 'staging.path' AND target_table = 'warehouse.fact_trip';

-- Update the watermark at the end of the run
INSERT INTO warehouse.etl_watermark (source_table, target_table, last_event_ts)
VALUES ('staging.path', 'warehouse.fact_trip', :max_event_ts)
ON CONFLICT (source_table, target_table)
DO UPDATE SET last_event_ts = EXCLUDED.last_event_ts;
\end{lstlisting}

\subsection{Idempotency and replay}
Idempotency is guaranteed by an \texttt{INSERT~\dots~ON~CONFLICT~DO~UPDATE} pattern on the natural key of every fact and dimension. A replay of the same window therefore produces the same final state, regardless of the number of repetitions. This property is essential for the binomial collaboration mode: either student can re-trigger a run on the shared Azure VM without risking divergence between the two members' working copies of the warehouse.

\section{Cleaning rule engine}

\subsection{Rule catalogue}
Eleven cleaning rules, identified \texttt{C1}--\texttt{C11}, are formalized at the entry of the warehouse layer. The rules are grouped into two families: \texttt{C1}--\texttt{C7} apply to the trip-related sources, and \texttt{C8}--\texttt{C11} apply to the high-resolution telemetry source. Each rule specifies a name, a severity level (critical, high, medium, low), a corrective action (reject, nullify, clamp) and a logical condition. Table~\ref{tab:rules-prep} reproduces the catalogue.

\begin{table}[h]
\centering
\caption{Cleaning rule catalogue applied at the warehouse boundary.}
\label{tab:rules-prep}
\begin{tabular}{|l|l|l|l|p{5.4cm}|}
\hline
\textbf{ID} & \textbf{Name} & \textbf{Sev.} & \textbf{Action} & \textbf{Condition} \\
\hline
C1 & temporal\_epoch\_error\_filter & critical & reject & \texttt{begin\_path\_time} $<$ 2019-10-01 \\
C2 & trip\_duration\_positive & critical & reject & \texttt{path\_duration} $\le 0$ \\
C3 & trip\_distance\_positive & high & reject & \texttt{distance\_driven} $\le 0$ \\
C4 & fuel\_used\_sanitize & critical & nullify & \texttt{fuel\_used} $\notin [0, 500]$ \\
C5 & speed\_cap & high & clamp & \texttt{max\_speed} $>$ 200 \\
C6 & stop\_duration\_bounds & medium & reject & duration $\le 0$ or $> 1$ year \\
C7 & device\_vehicle\_tenant\_chain & high & reject & device without vehicle link \\
C8 & archive\_ignition\_on\_for\_events & high & reject & \texttt{ignition} $\ne 1$ \\
C9 & archive\_accelerometer\_int8\_sanity & medium & reject & $|x|, |y|, |z| > 127$ \\
C10 & archive\_speed\_clamp & medium & clamp & speed $\notin [0, 250]$ \\
C11 & archive\_rpm\_clamp & low & clamp & rpm $\notin [0, 8000]$ \\
\hline
\end{tabular}
\end{table}

\subsection{Implementation pattern}
Each rule is implemented through an explicit SQL clause executed during the load of the corresponding fact table. Rejected rows are sent to a quarantine table \texttt{warehouse.quarantine\_rejected}, with the rule identifier, the rejection timestamp and the offending row payload. Listing~\ref{lst:rule-c2} shows the implementation of rule \texttt{C2}.

\begin{lstlisting}[style=sql,caption={Implementation of cleaning rule \texttt{C2} (positive trip duration).},label={lst:rule-c2}]
-- C2: reject trips with non-positive duration
WITH cleaned AS (
    SELECT *
    FROM staging.path
    WHERE path_duration > 0
),
rejected AS (
    INSERT INTO warehouse.quarantine_rejected (rule_id, rejected_at, payload)
    SELECT 'C2', NOW(), to_jsonb(p)
    FROM staging.path p
    WHERE p.path_duration <= 0
    RETURNING 1
)
SELECT COUNT(*) FROM cleaned;
\end{lstlisting}

\subsection{Quarantine table}
The quarantine table is structured as \texttt{(rule\_id, rejected\_at, payload~JSONB)}. The \texttt{JSONB} payload preserves the full content of the offending row, allowing the audit team to reconstruct the violation context without accessing the staging layer. The choice of \texttt{JSONB} is significant: it makes use of an idiomatic PostgreSQL feature that has no direct equivalent in the operational MySQL setup of \hostorg, and it illustrates the analytical leverage that motivated the migration in the first place.

\section{Dimensional modeling of the warehouse}

\subsection{Dimensions}
The warehouse layer hosts five dimensions and one bridge:

\begin{itemize}
    \item \texttt{dim\_tenant} --- billing entities, derived from the \texttt{user\_xxxxx} master data;
    \item \texttt{dim\_vehicle} --- vehicles, with normalized manufacturer and class;
    \item \texttt{dim\_device} --- GPS devices, restricted to the 616 devices linked to a vehicle;
    \item \texttt{dim\_driver} --- drivers (294 individuals);
    \item \texttt{dim\_date} --- calendar from 2019-10-01;
    \item \texttt{bridge\_device\_driver} --- SCD~Type~2 mapping between devices and drivers.
\end{itemize}

\subsection{Facts}
Eleven facts are produced from the cleaned staging data, summarized in Table~\ref{tab:facts-prep}.

\begin{table}[h]
\centering
\caption{Fact tables produced by the warehouse layer.}
\label{tab:facts-prep}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Fact table} & \textbf{Grain} & \textbf{Source} \\
\hline
\texttt{fact\_trip} & (tenant, device, begin\_time) & \texttt{staging.path} \\
\texttt{fact\_overspeed} & (tenant, device, begin\_time) & \texttt{staging.rep\_overspeed} \\
\texttt{fact\_stop} & (tenant, device, stop\_start) & \texttt{staging.stop} \\
\texttt{fact\_speed\_notification} & (tenant, device, created\_at) & \texttt{staging.notification} \\
\texttt{fact\_daily\_activity} & (tenant, device, day) & \texttt{staging.rep\_activity\_daily} \\
\texttt{fact\_harsh\_event} & (tenant, device, event\_ts) & \texttt{staging.archive} \\
\texttt{fact\_telemetry\_daily} & (tenant, device, date) & \texttt{staging.archive} \\
\texttt{fact\_maintenance} & (tenant, id\_maintenance) & \texttt{staging.maintenance} \\
\texttt{fact\_fueling} & (tenant, doc\_id) & fueling sources \\
\hline
\end{tabular}
\end{table}

\subsection{UPSERT pattern}
The loading of every fact relies on the canonical PostgreSQL \texttt{INSERT~\dots~ON~CONFLICT~DO~UPDATE} pattern shown in Listing~\ref{lst:upsert}.

\begin{lstlisting}[style=sql,caption={Idempotent upsert into \texttt{fact\_trip}.},label={lst:upsert}]
INSERT INTO warehouse.fact_trip (
    tenant_id, device_id, begin_path_time,
    end_path_time, distance_km, max_speed_kmh, duration_seconds
)
SELECT tenant_id, device_id, begin_path_time, end_path_time,
       distance_driven, LEAST(max_speed, 200) AS max_speed_kmh,
       path_duration
FROM staging.path
WHERE begin_path_time > :watermark
  AND path_duration > 0
  AND distance_driven > 0
  AND begin_path_time >= '2019-10-01'
ON CONFLICT (tenant_id, device_id, begin_path_time) DO UPDATE
SET end_path_time   = EXCLUDED.end_path_time,
    distance_km     = EXCLUDED.distance_km,
    max_speed_kmh   = EXCLUDED.max_speed_kmh,
    duration_seconds= EXCLUDED.duration_seconds;
\end{lstlisting}

\section{Analytical marts}

\subsection{Behaviour mart}
The mart \texttt{marts.mart\_device\_monthly\_behavior} carries the 35 behavioural features at the (\texttt{tenant\_id}, \texttt{device\_id}, \texttt{year\_month}) grain and is rebuilt incrementally from the trip-related facts and the archive-derived facts. Its schema is the principal contract of the ML serving layer.

\subsection{Telemetry mart}
The mart \texttt{marts.mart\_device\_monthly\_telemetry} is the archive-derived companion of the behaviour mart, with engine-on time, idle ratio, average and high RPM indicators, computed from the high-resolution archive table.

\subsection{BI marts}
Three additional marts feed the BI layer: \texttt{mart\_fleet\_daily}, \texttt{mart\_vehicle\_monthly} and \texttt{mart\_tenant\_monthly\_summary}. They are exposed through dedicated views (\texttt{v\_executive\_dashboard}, \texttt{v\_operational\_dashboard}, \texttt{v\_maintenance\_dashboard}) consumed by the BI dashboard.

\subsection{Unified ML view}
The view \texttt{v\_ml\_features\_full} unifies the behaviour and the telemetry marts into a single contract for the modeling notebooks; the companion view \texttt{v\_device\_risk\_profile} computes the weighted risk score and its categorical band.

\section{Feature engineering}

\subsection{Feature registry}
The catalogue of features is declared in \texttt{config/feature\_definitions.yaml}. Each feature is described by its name, its type, its unit, its formula and the fact tables it depends on. A typed Python wrapper exposes the registry to the notebooks. Table~\ref{tab:features-sample} shows a sample of the registry.

\begin{table}[h]
\centering
\caption{Sample of the 35-feature registry (selected groups).}
\label{tab:features-sample}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Group} & \textbf{Feature} & \textbf{Definition} \\
\hline
Volume & \texttt{total\_trips} & Count of trips in month \\
Volume & \texttt{total\_distance\_km} & Sum of distances \\
Volume & \texttt{stddev\_trip\_distance} & Std.\ dev.\ of trip distances \\
Speed & \texttt{avg\_max\_speed\_kmh} & Mean of per-trip max speed \\
Speed & \texttt{p95\_max\_speed} & 95\textsuperscript{th} percentile of max speed \\
Speed & \texttt{high\_speed\_trip\_ratio} & Trips with max speed $>$ 100 \\
Overspeed & \texttt{overspeed\_per\_100km} & Events / total\_distance $\times$ 100 \\
Harsh & \texttt{harsh\_brake\_per\_100km} & Brake events / 100~km \\
Harsh & \texttt{harsh\_accel\_per\_100km} & Acceleration events / 100~km \\
Temporal & \texttt{night\_trip\_ratio} & Trips between 20:00--06:00 \\
Temporal & \texttt{rush\_hour\_trip\_ratio} & Trips during peak hours \\
Telemetry & \texttt{monthly\_idle\_ratio} & Idle / engine-on minutes \\
Telemetry & \texttt{avg\_rpm} & Mean RPM across day \\
\hline
\end{tabular}
\end{table}

\subsection{Aggregation strategy}
The aggregation strategy follows three principles: \emph{normalization by exposure} (most counters are divided by 100~km of distance), \emph{operational interpretability} (each feature corresponds to an indicator that a fleet manager recognizes) and \emph{declarative specification} (the registry is the single source of truth between the SQL transformations and the Python feature consumption).

\subsection{Transformations}
The categorical features (vehicle class, manufacturer family) are encoded with a one-hot scheme; the ordinal features (vehicle class) are encoded with an explicit ordinal scheme; the quantitative features used by the unsupervised models are standardized per tenant by means of \texttt{StandardScaler}. An exploratory supervised track applies SMOTE to neutralize an anticipated class imbalance.

\section{Validation of the prepared dataset}

\subsection{Validation suite design}
The validation suite is implemented in \texttt{sql/99\_validation\_suite.sql} and gathers eight check categories \texttt{V1}--\texttt{V8}: dimension population (V1), referential integrity (V2), enforcement of \texttt{C1} (V3), enforcement of \texttt{C2}--\texttt{C3} (V4), enforcement of \texttt{C5} (V5), null-rate budgets on critical columns (V6), mart grain coverage (V7) and risk score distribution sanity (V8).

\subsection{Final dataset summary}
At the conclusion of the data preparation phase, the modeling-ready dataset comprises approximately \textbf{1{,}723 device-month observations} on the period 2025-01--2026-04, after applying the threshold of 100~km of total distance and a minimum number of trips per device-month. This dataset feeds the modeling experiments of Chapter~\ref{chap:model}.

\section{Conclusion}
This chapter has described the full preparation chain --- from the conversion of the MySQL dump files of \hostorg{} into PostgreSQL-importable artefacts, through the watermark-driven ETL pipeline and the eleven-rule cleaning engine, to the dimensional modeling of the warehouse, the materialization of the analytical marts, the engineering of the feature registry and the validation of the prepared dataset. The next chapter exploits the ML-ready dataset to train and to evaluate the unsupervised models that produce the behavioural segmentation and the risk score.
