% =====================================================================
%  Acknowledgments
% =====================================================================
\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
\thispagestyle{plain}

The work documented in this thesis is the result of six intense months during which many people invested time, expertise and goodwill on my behalf. It is a privilege to thank them here.

I am profoundly grateful to my academic supervisor, \supervisor, for the rigour of his guidance, the precision of his remarks and the patience with which he examined every iteration of the deliverables. His insistence on substance over form has meaningfully shaped the engineering posture of this project.

I extend my warmest thanks to my industry supervisor, \industrysupervisor, who welcomed me at \hostorg, granted me access to the operational telematics platform and trusted me with a refactoring scope of significant impact. The technical conversations we had on the warehouse design and on the production pipeline have considerably enriched the present work.

My sincere thanks go to the entire engineering team of \hostorg{} who answered my questions on the data, the operational regimes of the fleet and the constraints of the Microsoft Azure environment on which the platform is deployed. The remote collaboration through SSH tunnelling and the controlled access to the cloud database would not have been as smooth without their day-to-day availability.

I would like to thank the members of the jury who have honoured this defence with their attention and whose remarks I will study carefully.

Finally, my deepest gratitude goes to my family and to my close friends, whose constant support has been the silent infrastructure of this entire project.

\cleardoublepage
